---
title: "Assignment 7"
author: "Sofie Pauline Rise"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)  
library(rvest)    
library(lubridate)
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)
```


##### In this assignment we will look at data from new borns in the U.S from 2017-2019
First we will look at the portion of boys to girls in the given years.
Then we will look at the average weight in grams by gender. 
Last we will look at the portions of boys to girls by day of birth 

```{r,echo=TRUE, message = FALSE, warning= FALSE, include=FALSE}
##loading data
baby2017<- "Nat2017PublicUS.c20180516.r20180808.txt"
baby2018<- "Nat2018PublicUS.c20190509.r20190717.txt"
baby2019<- "Nat2019PublicUS.c20200506.r20200915.txt"

##selecting columns and give name to them 
data2017 <- 
  read_fwf(baby2017,
           fwf_positions(start = c(9,13, 23, 475, 504),
                         end   = c(12,14, 24, 476, 507)))

data2018 <- 
  read_fwf(baby2018,
           fwf_positions(start = c(9,13, 23, 475, 504),
                         end   = c(12,14, 24, 476, 507)))


data2019 <- 
  read_fwf(baby2019,
           fwf_positions(start = c(9,13, 23, 475, 504),
                         end   = c(12,14, 24, 476, 507)))
      
##make csv file
write.csv(data2017, "birth2017.csv")
write.csv(data2018, "birth2018.csv")
write.csv(data2019, "birth2019.csv")

csv.2017<- read_csv("birth2017.csv")
csv.2018<- read_csv("birth2018.csv")
csv.2019<- read_csv("birth2019.csv")

##combining all the csv togheter
total<- rbind(csv.2017,csv.2018,csv.2019)
##changing column names 
colnames(total) <- c("Number","Year","Month","Weekday", "Gender", "Weight")
##making the weight variable to numeric 
total$Weight<- as.numeric(total$Weight)

##start to separate the variables to get ready to plot
##first make a data frame for each gender and each year 
g <- total %>% 
  filter(Gender=="M")
  
 
##same for girls
j<- total %>%
  filter(Gender=="F")
##assignment 1, portion of boys to girls 
##looking at all years combined so 2017-2019


boy<- g %>%
  group_by(Year)%>%
  tally() %>%
  mutate(Gender="M")

girl<-j %>%
  group_by(Year)%>%
  tally()%>%
  mutate(Gender= "F")
##combining the two data frames to one 
## three variables, "Year" and "n" for number of births, and gender 
bandg<- rbind(boy,girl)

##cahnging variable Year to character to easier be able to plot
bandg$Year <- as.character(bandg$Year)
```


```{r,message = FALSE, warrning= FALSE,echo=FALSE}
##plotting
##selectiong Gender to x and n(number) to y
##geombar is box/bar outlay 
##selecting to separate the bars i three cathegoriez based on year

bandg%>%
  ggplot(aes(x=Gender, y=n, fill=Gender))+
  geom_bar(stat = "identity")+
  facet_grid(~Year)+
  ggtitle("The proportion of boys to girls")+
  ylab("Births")
```

##### As we can see here there are more boys born in the U.S during the period of 2017-2019. 
```{r,echo=TRUE, message = FALSE, warning= FALSE, include= FALSE}
## As we can see in all three years there have been more baby boy births 


##2 averange birth weight 

##finding the avg. weight and grouping by year 
##biys first
avgB<- g%>%
  group_by(Year)%>%
  mutate(Average=mean(Weight))
##girls
avgG<- j%>%
  group_by(Year)%>%
  mutate(Average=mean(Weight))

##filtering by year to find a number for each year 
##selecting what columns to look at here, gender, year and the average
## then slice to make it one observation (mean of all months and days)
##2017
avgG17 <- avgG %>% 
  filter(Year==2017) %>% 
  select(Gender, Year, Average)%>%
  slice(1)
  
#2018
avgG18 <- avgG %>% 
  filter(Year==2018) %>% 
  select(Gender, Year, Average) %>% 
  slice(1)
#2019
avgG19 <- avgG %>% 
  filter(Year==2019) %>% 
  select(Gender, Year, Average) %>% 
  slice(1)

##doing the same for boys 
#2017
avgB17 <- avgB %>% 
  filter(Year==2017) %>% 
  select(Gender, Year, Average) %>% 
  slice(1)

#2018
avgB18 <- avgB %>% 
  filter(Year==2018) %>% 
  select(Gender, Year, Average) %>% 
  slice(1)

avgB19 <- avgB %>% 
  filter(Year==2019) %>% 
  select(Gender, Year, Average) %>% 
  slice(1)

##to be able to plot make one data from all the average weight from all the years
df.mean<- rbind(avgB17,avgB18,avgB19,avgG17,avgG18,avgG19)


```

```{r,message = FALSE, warrning= FALSE,echo=FALSE}
##plotting of average weight
##gender on the x and Average weight on the y, and sep. by gender
##telling it to be a bar graph 

df.mean%>%
  ggplot(aes(x=Gender, y=Average, fill=Gender))+
  geom_bar(stat = "identity")+
  facet_grid(~Year)+
  ggtitle("The average weight of new borns in grams", 
          subtitle = "Separated by gender 2017-2019")

```
##### Here we can see that the boys are a bit heavier than girls at birth.


```{r,echo=TRUE, message = FALSE, warning= FALSE, include=FALSE}
##3
##portion of boys to girls by day of birth 
##How many boys and girls were born on each day? 
##first for girls, using the old data wich is grouped by year 
##selecting only F to be in this group 
##doing for each year first and then combining to one data frame in the end

#2017
boyday17<- csv.2017%>%
  filter(X4=="M")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="M")%>%
  mutate(Year="2017")

girlday17<- csv.2017%>%
  filter(X4=="F")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="F")%>%
  mutate(Year="2017")


##same for 2018 
girlday18<- csv.2018%>%
  filter(X4=="F")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="F")%>%
  mutate(Year="2018")

boyday18<- csv.2018%>%
  filter(X4=="M")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="M")%>%
  mutate(Year="2018")



##2019
boyday19<- csv.2019%>%
  filter(X4=="M")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="M")%>%
  mutate(Year="2019")

girlday19<- csv.2019%>%
  filter(X4=="F")%>%
  group_by(X3)%>%
  tally()%>%
  mutate(Gender="F")%>%
  mutate(Year="2019")




##combining to one data
day.t<- rbind(boyday17,boyday18,boyday19,girlday17,girlday18,girlday19)
colnames(day.t)<- c("Weekday", "Number", "Gender", "Year")



```

```{r,message = FALSE, warrning= FALSE,echo=FALSE}
day.t%>%
  ggplot(aes(x=Gender, y=Number, fill=Year))+
  geom_bar(stat = "identity")+
  facet_grid(~Weekday)+
  ggtitle("The proportion of boys to girls by day of birth ", 
          subtitle = "Where Sunday is first day of the week")+
  ylab("Number of births")
```
##### Here we can see that most child are born in Monday-Friday for both genders 
